### 2021-02-12 Modules/sec_prlm_localization.tex
Sent as draft No.2

Following the classical methods overview in \cite{krim1996two},most localization estimators may be divided to two main groups - i.e. \textit{spectral-based} and \textit{parametric}.
The basic difference between the two groups is the assumption of impinging signal's statistical model when using parametric approach, which leads to higher accuracy in the expense of computation effort.
In both approaches, the estimations are based on calculation of several hypothesis and choosing the peaks in the generated graph.
Considering the search for a single parameter (e.g. DOA in a 2D plane), the two approaches are pretty similar in the sense that in both, the processor computes the confidence in a predetermined set of hypothesis which is followed by a single dimension scan.
The difference is more evident when considering multiple parameter, such as spherical direction (i.e. azimuth and elevation) or even actual location adding also the distance to the target of interest.
In the spectral based, each parameter is estimated independently of the other parameters, resulting in several 1D scans (as the number of estimated parameters), where the parametric approach will result in a multi-dimensional search for the best fitted parameters set.
In the following, a basic overview of some classical algorithms is presented.
The earliest localization algorithms, which date back to world war II, are the beamforming based, e.g. the conventional Bartlet beamformer.
These methods, although appealing in terms of complexity and computation effort, suffer from fundamental limitation, e.g. its spatial performance depend solely on the array aperture and no improvement is achieved when more data is collected.
Another class of estimators, where a statistical model is assumed (which is also the drawback of those methods), is the parametric estimators such as Maximum-Likelihood \cite{TWODECADES-81/106} and Maximum-Entropy \cite{TWODECADES-23}.

@@

Trying to relax the need for such accuracy, also serving as the origin of the sub-space based approach, was the Multiple SIgnal Classification (MUSIC) \cite{MUSIC_} algorithm.
In general, the MUSIC algorithm consists of several stages:
    \item Estimation of the covariance matrix
    \item Eigen decomposition - extracting the eigen values and its related eigen-vectors
    \item Estimating the number of sources
    \item Projecting the samples over the array manifold
    \item Finding the largest peaks according to the estimated number of sources

@@

Another, probably even more important breakthrough, was the introduction of the array manifold and noise space (hence its name - sub-space) which proved to be orthogonal, thus allowing the use of orthogonal projection.
In the very thorough review of classical algorithms \cite{TWODECADES}, two approaches - i.e. the parametric approach and ULA-specific algorithms are also discussed.
the parametric approach, where stochastic model is assumed and Maximum Likelihood (ML) and Stochastic Maximum Likelihood (SML) are employed, is less relevant to this work therefore not discussed.
In the set of ULA-specific algorithms, including ROOT-MUSIC \cite{ROOT_MUSIC}, ESPRIT (Estimation of Signal Parameters via Rotational Invariance Techniques) \cite{ESPRIT} and some iterative methods.
We choose to emphasize on the ESPRIT method, being an important milestone in the path to state-of-the-art algorithms.
In addition to using the rotational invariance of the signal sub-space eigenvectors, it also reduced computation and storage costs (especially in the multi-dimensional estimation case) by replacing the covariance matrix calculation and eigendecomposition by a relaxed partial singular value decomposition (SVD) which is employed on the data itself without squaring it - mitigating numerical problems associated with ill-conditioned matrices.

@@

    \item High Order Statistics (HOS, also refereed as comulants), which extracts more information from the samples' higher order moments, is a very active field of research due to some fundamental issues which are inherently resolved - i.e. the Gaussian noise vanishes in the 4th order statistics and the ability to resolve more DOAs than array elements.

@@

    \item Sparsity based estimation, using non-Frobenius norm used in the algorithms development, has proved to improve estimation resolution in some cases \cite{LocalizationOfMultipleSpeaker}, namely high reverbrant acoustic scenarios, but is less relevant to this work therefore not elaborated.

@@

    It is also less relevant to this work.
    \item Naturally, also many trials of harnessing the promising concept of neural networks are being done, for example \cite{NNDOA}

@@

In this work, we actually revisit the most basic approach - i.e. beamforming.

>>>>

Following the classical methods overview in \cite{krim1996two},localization estimators may be divided to two main groups - i.e. \textit{spectral-based} and \textit{parametric}.
The spectral based estimators steer the array (mechanically or by beamforming) to a set of DOAs, searching for peaks in the received energy, where each peak is treated as a resolved emitter when greater than a certain threshold.
This approach of spatial filtering, is considered to suffers from fundamental
limitations, namely ``its performance, is directly dependent upon the physical size of the array (the aperture), regardless of the available data collection time and signal-to-noise ratio (SNR)`` \cite{krim1996two}.
Although trials were made to increase the resolution of an array given a certain aperture, this notion remains intact to these days.
In this contribution we tackle this statement as will shown in Chapter\ref{chap:firstchap}.
The later approach, of parameter estimation assumes a physical model of the experimental scenario, which serve a basis for the development of spatio-temporal estimators which takes advantage not only from the spatial diversity but also from the temporal information collected via sequential time separated snapshots in order to increase the overall SNR.
Such methods, not only enabled the increase of spatial filtering resolution, but also allowed the implementation of data processing algorithms, more sophisticated than a mere search with a steered array.
In the following, a basic overview of some classical algorithms are presented.
The earliest localization algorithms, which date back to world war II, are the beamforming based, e.g. the conventional Bartlet \cite{van2004optimum} and Capon's \cite{capon1969high} beamformers.
Early examples of parametric estimators, are the Maximum-Likelihood \cite{macdonald1969optimum,schweppe1968sensor} and Maximum-Entropy \cite{ables1974maximum} which assume a known PDF to the received signal and estimate the desired parameters according to empirical results and fitting the closest matching PDF.

@@

Trying to relax the need for such accuracy, also serving as the origin of the sub-space based approach, was the Multiple SIgnal Classification (MUSIC) \cite{schmidt1986multiple} algorithm.

@@

Another, probably even more important breakthrough, was the introduction of the orthogonal array manifold noise spaces, thus allowing the use of orthogonal projection in order to mitigate the noise effects.
Another important parametric approach related algorithm, is the ESPRIT (Estimation of Signal Parameters via Rotational Invariance Techniques) \cite{ESPRIT}, serving as a milestone in the path to state-of-the-art algorithms.
In addition to using the rotational invariance of the signal sub-space eigenvectors, it also reduced computation and storage costs (especially in the multi-dimensional estimation case) by replacing the covariance matrix calculation and eigendecomposition by a relaxed partial singular value decomposition (SVD) which is employed on the data itself without squaring it - thus mitigating numerical problems associated with ill-conditioned matrices.

@@

    \item High Order Statistics (HOS, also refereed as cumulants), which extracts more information from the samples' higher order moments, is a very active field of research due to some fundamental issues which are inherently resolved - i.e. the Gaussian noise vanishes in the 4th order statistics and the ability to resolve more DOAs than array elements \cite{chevalier2006high}.

@@

    \item Inspired by the vast research field of sparse representations some algorithms \cite{nadiri2014localization} use non-Frobenius norm (i.e. $L^{p}$ norms where $p<2$), in order improve estimation resolution under high reverberation acoustic scenarios.

@@

    \item Naturally, also many trials of harnessing the promising concept of neural networks are being done, for example \cite{shareef2008localization}.

@@

In this work, we actually revisit the most basic approach - i.e. beamforming which resides under the spectral based algorithms.

### 2021-02-12 Modules/sec_prlm_propWaveField.tex
Sent as draft No.2

Where $A$ is a complex constant, $\omega$ denotes a constant temporal frequency and $k_{x}x,k_{y},k_{z}$ are real constants. 

@@

>>>>

Where $A$ is a complex constant, $\omega$ denotes temporal radial frequency and $k_{x}x,k_{y},k_{z}$ are real constants. 

@@

\par For example, in the context of ULA, assuming far field scenario, the impinging waves are treated as constant phase planes as in \eqref{eq_prlm_waveEq_finSol}.
The spatial diversity of the ULA elements is expressed as a TOA difference of the impinging signal in each sensor which is will be shown to provide clues for the signal's DOA as can be seen in Fig.~\ref{fig_ULA_sketch}.
        \begin{overpic}[width=0.6\linewidth, 
        tics=10,trim=0 0 0 0]{./Media/arraySketch.png}
        \put(20,45){\rotatebox{-50}{\tiny{Impinging wavefront}}}
        \put(33.25,2.25){\rotatebox{0}{\tiny{$d$}}}
        \put(52.25,26){\rotatebox{38}{\tiny{$c\tau=d\cos{\theta}$}}}
        \put(49,15.75){\rotatebox{0}{$\theta$}}
        \put(15,7){\rotatebox{0}{\tiny{0}}}
        \put(28,7){\rotatebox{0}{\tiny{1}}}
        \put(41,7){\rotatebox{0}{\tiny{2}}}
        \put(80,7){\rotatebox{0}{\tiny{$N-1$}}}
        \put(44,36){\rotatebox{38}{\tiny{$t=\rBrace{N-1}\tau$}}}
        \put(57,36){\rotatebox{38}{\tiny{$t=\rBrace{N-2}\tau$}}}
        \put(70,36){\rotatebox{38}{\tiny{$t=\rBrace{N-3}\tau$}}}
        \put(92,25){\rotatebox{38}{\tiny{$t=0$}}}
     \caption{An illustration of an $N$ element ULA, impinged with a plane wave arriving from a DOA of $\theta$.}
    \label{fig_ULA_sketch}
When also considering the narrowband scenario, where TOA difference is merely phase shift, it seems obvious that by phase measuring, one should be able to extract the DOA from the spatially sampled data.

### 2021-02-12 Modules/sec_prlm_sensorArrays.tex
Sent as draft No.2

Microphone arrays consist of sets of microphones positioned in a way to function as a directional acoustic antenna. 
Microphone arrays are utilized for filtering signals in a space-time field. 
Filtering is enabled by exploitation of the incoming signals spatial characteristics. 
A desirable spatial filtering, i.e., beamforming, should result in enhancement of signals of interest, originated in a specific direction, while forcing suppression of undesired signals originated in other directions.
Microphone arrays are utilized for solving many signal processing problems: dereverberation, localization of a single source, noise reduction and source separation \cite{cohen2004multichannel,habets2006dual,pavlidi2013real}.
\par 
Numerous factors must be taken into consideration when designing a microphone array configuration.
Initially, the geometry of the microphone array plays an important role in the formulation of the processing algorithms, as it forces fundamental constraints on the array’s operation. 
In most cases, the array geometry is the first consideration in array design due to practical and physical constraints of the design. 
Therefore, the degree of freedom in choosing the array geometry is limited. 
Nevertheless, in some other crucial problems such as noise reduction or source separation, the geometry of the array may have little importance. 
For instance, Uniform Linear Arrays (ULA) can only handle one source direction at a time resulting in uncertainties and possibly direction ambiguities.
Therefore, ULA configurations must be ruled out when handling several signal sources.
Other array geometries such as non-uniform linear arrays and circular arrays, have been studied in the field \cite{liu2008design,van2004optimum}.
This work is based on ULA and therefore we elaborate on this geometry.
A conventional beamformer \cite{van2004optimum} is composed of a ULA of microphones. 
That is, the microphones are positioned on an axis with a uniform spacing between the microphones.
The general expression of the microphone positioning in a ULA is given by:
\begin{equation}
x_{m}=m\cdot{d}, m=0, 1, \dots, M-1
\end{equation}
where $m$ denotes the microphone index; $x_{m}$ denotes the position of the microphone having index $m$, where $x_{m} = 0$ corresponds to the left-hand side of the array, i.e., the location where the microphone indexed by $m = 0$ is positioned; $d$ denotes the spacing between microphones, and $M$ is the array size, i.e., the number of microphones in the array, as demonstrated in Fig.~\ref{fig_ULA}.
\begin{figure}[h!]
    \begin{center}
        \begin{overpic}[width=0.5\linewidth, 
        %grid, 
        tics=10,trim=0 0 0 0]{./Media/fig_ULA.png}
            % \put (50, 62.5) {\footnotesize{$r=0$}}
        \end{overpic}
    \end{center}
     \caption{A uniform linear array of size M and spacing d.}
    \label{fig_ULA}
\end{figure}

>>>>

Sensor arrays are sets of independently positioned sensors, where each sensor is sampling temporal snapshots of impinging signals.
The samples from the entire array are then fused together in order to extract the underlying data.
The spatial diversity of the sampled data allows the extraction of spatial features which are not accessible by mere temporal analysis of the data.
Sensor arrays are used in numerous applications, ranging from source localization, through communication, medical applications, astronomy etc.
\par 
When designing a sensor array, multiple consideration are to be taken into account, ranging from physical characteristics e.g. area, weight and carrying platform, through performance related considerations such as accuracy, spatial selectivity, signal to noise ratio and even the financial aspect.
Obviously, the most fundamental array specification is its geometry, for it dictates the spatial relations between simultaneous samples measured by the array's sensors, which also greatly influences the processing methods of the raw-data.
A basic example of sensor array is the ULA (see Fig.~\ref{fig_ULA}), where its elements are uniformly spaced on a straight line with distance $d$ between a sensor to its nearest neighbour.
Considering an array of $N$ elements, the sensors are positioned at
\begin{equation}
p_{n}=nd,\ n=0,\dots,N-1
\end{equation}
where $n$ is the sensors index; $p_{n}$ denotes the position of the $n$'th sensor,$p_{0} = 0$ corresponds to the left-hand side of the array.
\begin{figure}[h!]
    \begin{center}
        \begin{overpic}[width=0.5\linewidth, 
        %grid, 
        tics=10,trim=0 0 0 0]{./Media/arrayBasic.png}
        \put (4, 4.5) {\tiny{$N$}}
        \put (86, 4.5) {\tiny{$N-1$}}
        \put (83.25, 28.5) {\tiny{$\rBrace{N-1}\cdot{}d$}}
        \end{overpic}
    \end{center}
     \caption{A uniform linear array of size $N$ and spacing $d$.}
    \label{fig_ULA}
\end{figure}
\par To proper formulate the discussed additional spatial information when using sensor array based processors, a quick overview on wave propagation is due and presented in the following section (Sec.~\ref{sec:prlm_propWaveField}).

### 2021-02-12 Modules/sec_prlm_timeSpaceSig.tex
Sent as draft No.2

% The incoming signals are spatially sampled by the array microphones, then the samples are processed to attenuate signals from undesired directions and extract the signal from a desired direction. 
% A spatial response of the microphone array is obtained via a beam (main-lobe) directed to the desired signal while nulls are directed to the undesired signals.
As mentioned before, the measurements of the sensor array's elements are fused together, where spatial information resides in the relations between samples taken in the same time $t$.
A basic sensor array based entity is the beamformer, which is designed to enhance signals from certain directions while suppressing other signals from unwanted directions, heuristically forming a beam directed to the enhanced direction.  
\begin{figure}[ht!]
    \begin{center}
        \begin{overpic}[width=0.5\linewidth, 
        %grid, 
        tics=10,trim=0 0 0 0]{./Media/fig_ULA_imping.png}
            % \put (50, 62.5) {\footnotesize{$r=0$}}
        \end{overpic}
    \end{center}
     \caption{A simple beamformer design.}
    \label{fig_ULA_imping}
\end{figure}
Fig.~\ref{fig_ULA_imping} illustrates a beamformer design based on a ULA of Fig.~\ref{fig_ULA}.
% Let $\vecnot{f}\rBrace{t,\vecnot{x}}$ denote the set of signals sampled by the microphone array at time $t$, expressed by
A temporal snapshot measured by the array at time $t$ is
\begin{equation}
\vecnot{f}\rBrace{t,\vecnot{x}} = \vBrace{f\rBrace{t,x_{0}},\dots,f\rBrace{t,x_{M-1}}}^{T},
\end{equation}
where $x_{m}$ denotes the microphone position, $t$ denotes the continuous-time variable and $^{T}$ denotes the transpose operation. 
Assuming simple, weight-and-sum beamformer, the output ($y\rBrace{t}$) is expressed by:
\begin{equation}
y\rBrace{t} = \sum_{m=0}^{M-1}{f\rBrace{t,x_{m}}w_{m}^{*}},
\end{equation}
where $w_{m}$ is a complex weight used for the $m$'th sensor's output, as shown in Fig.~\ref{fig_ULA_imping} and $^{*}$ denotes complex conjugation. 
Consider $s\rBrace{t} = e^{j\omega{t}}$ as a plane wave propagating at angular frequency $w$, and $\theta \in \vBrace{-\frac{\pi}{2},\frac{\pi}{2}}$ as the direction of arrival (DOA) angle measured with respect to the broadside of the linear array, as shown in Fig.~\ref{fig_ULA_imping}.
The wave signals spatially sampled by the sensor array inputs are
\begin{equation}
\label{eq_prlm_timeSpace_outputVec}
\vecnot{f}\rBrace{t,x} = \vBrace{s\rBrace{t-\tau_{0}}, \dots, s\rBrace{t-\tau_{M-1}}}^{T}, \tau_{m} = \frac{\sin{\theta}\cdot{}x_{m}}{c},
\end{equation}
where $\tau_{m}$ is the propagation delay for the incoming signal and $c$ is the wave's velocity in the medium.
% A value of $\tau_{m}=0, \forall{m}$ implies a DOA of $\theta = 0$, i.e. a plane wave parallel to the array, propagating perpendicularly to the array.
Let $\kappa=\frac{\omega}{c}\sin{\theta}=\frac{2\pi}{\lambda}\sin{\theta}$ denote the wavenumber for plane waves in a locally homogeneous medium, where $\lambda$ denotes the wavelength corresponding to the angular frequency $\vecnot{d}\rBrace{\kappa}$ denote the \emph{steering vector}, featuring all of the array's spatial characteristics.
Based on \eqref{eq_prlm_timeSpace_outputVec}, and the definition of $\kappa$ above, the steering vector can be expressed as
\begin{equation}
\vecnot{d}\rBrace{\kappa}=\vBrace{e^{-j\kappa{}x_{0}},\dots,e^{-j\kappa{}x_{M-1}}}^{T}.
\end{equation}

>>>>

As mentioned before, the measurements of the sensor array's elements are fused together, where spatial information resides in the relations between samples taken the same time $t$.
A basic application of the sensor array is the beamformer, designed to enhance signals from certain directions while suppressing other signals from unwanted directions, forming a beam directed to the enhanced direction.  
\begin{figure}[ht!]
    \begin{center}
        \begin{overpic}[width=0.6\linewidth, 
        %grid, 
        tics=10,trim=0 0 0 0]{./Media/arrayProc.png}
        \put(20,65){\rotatebox{-50}{\tiny{Impinging wavefront}}}
        \put(49,33.25){\rotatebox{0}{$\theta$}}
        \put(15,25){\rotatebox{0}{\tiny{0}}}
        \put(28,25){\rotatebox{0}{\tiny{1}}}
        \put(41,25){\rotatebox{0}{\tiny{2}}}
        \put(80,25){\rotatebox{0}{\tiny{$N-1$}}}
        \put(16,20){\rotatebox{0}{\tiny{$\omega_{0}$}}}
        \put(29.5,20){\rotatebox{0}{\tiny{$\omega_{1}$}}}
        \put(42,20){\rotatebox{0}{\tiny{$\omega_{2}$}}}
        \put(81,20){\rotatebox{0}{\tiny{$\omega_{N-1}$}}}
        \put(50.5,3.5){\rotatebox{0}{\tiny{$z\rBrace{t}$}}}
        \end{overpic}
    \end{center}
     \caption{ULA based coherent processing of simultaneous spatially sampled signal, measuring a wavefront impinging from DOA $\theta$}
    \label{fig_ULA_imping}
\end{figure}
In complimentary to Fig.~\ref{fig_ULA_sketch}, Fig.~\ref{fig_ULA_imping} adds the processor layer of the beamformer.
\par A temporal snapshot measured by the array at time $t$ is
\begin{equation}
\vecnot{f}\rBrace{t,\vecnot{p}} = \vBrace{f\rBrace{t,p_{0}},\dots,f\rBrace{t,p_{N-1}}}^{T},
\end{equation}
where $p_{n}$ denotes the $n$-th sensor position, $t$ denotes the continuous-time variable and $^{T}$ denotes the transpose operation. 
Considering the \emph{conventional beamformer} \cite{van2004optimum}, the output ($y\rBrace{t}$) is expressed by:
\begin{equation}
z\rBrace{t} = \sum_{n=0}^{N-1}{f\rBrace{t,p_{n}}\omega_{n}^{*}},
\end{equation}
where $\omega_{n}$ is a complex weight used for the $n$'th sensor's output, as shown in Fig.~\ref{fig_ULA_imping} and $^{*}$ denotes complex conjugation. 
Consider $s\rBrace{t} = e^{j\omega{t}}$ as a plane wave propagating at angular frequency $w$, and $\theta \in \vBrace{-\frac{\pi}{2},\frac{\pi}{2}}$ as the direction of arrival (DOA) angle measured with respect to the broadside of the linear array, as shown in Fig.~\ref{fig_ULA_imping}.
The wave signals spatially sampled by the sensor array inputs are
\begin{equation}
\label{eq_prlm_timeSpace_outputVec}
\vecnot{f}\rBrace{t,x} = \vBrace{s\rBrace{t},s\rBrace{t-\tau}\dots, s\rBrace{t-\rBrace{N-1}\tau}}^{T},
\end{equation}
where $\tau = \frac{d\cdot{}\cos{\theta}}{c}$ is the propagation delay between consecutive sensors and $c$ is the wave's velocity in the medium.
Considering narrowband signals (i.e. $s\rBrace{t}=\exp^{-j\omega{}t}$), the time delay $\tau$ translates to a mere phase shift and an auxiliary dephased \emph{steering vector} takes the form of
\begin{equation}
\vecnot{d}\rBrace{\theta}=\vBrace{1,e^{-j\omega\tau},\dots,e^{-j\rBrace{N-1}\omega\tau}}^{T}.
\end{equation}
As every impinging signal will cause the excitation of the array elements in the form of a specific steering vector (up to a common additive phase), we define the \emph{array manifold} to be the set of all steering vectors of $\theta\in\left[0,2\pi\right)$, hence it also spans the received signal sub space in the noiseless scenario.
